---
title: "03 Exploratory analysis"
author: "Alexandra Tejada-Strop"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Load Data
```{r}
library(tidyverse)
library(here)
library(skimr)
library(scales)

dir.create(here("results", "tables"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("results", "figures"), showWarnings = FALSE, recursive = TRUE)

analytic_dat <- readRDS(here("data", "processed-data", "prams_analytic.rds"))

dim(analytic_dat)
```

## Screening distribution
```{r}
screen_tbl <- analytic_dat |> 
  count(screened) |> 
  mutate(
    pct = n / sum(n),
    screening_status = case_when(
      screened == 0 ~ "Not screened",
      screened == 1 ~ "Screened",
      TRUE ~ as.character(screened)
    )
  )

screen_tbl

saveRDS(
  screen_tbl,
  here("results", "tables", "screening_distribution.rds")
)
```

## Postpartum depression distribution
```{r}

ppd_tbl <- analytic_dat |> 
  count(pp_depress_bin) |> 
  mutate(
    pct = n / sum(n),
    depression_status = case_when(
      pp_depress_bin == 0 ~ "No depressive symptoms",
      pp_depress_bin == 1 ~ "Postpartum depressive symptoms",
      TRUE ~ as.character(pp_depress_bin)
    )
  )

ppd_tbl

```

## Plot: Postpartum depressive symptoms by screening

```{r}
data_summary <- analytic_dat |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(proportion = n / sum(n)) |>
  mutate(
    screening_status = case_when(
      screened == 0 ~ "Not screened",
      screened == 1 ~ "Screened",
      TRUE ~ as.character(screened)
    ),
    depression_status = case_when(
      pp_depress_bin == 0 ~ "No depressive symptoms",
      pp_depress_bin == 1 ~ "Postpartum depressive symptoms",
      TRUE ~ as.character(pp_depress_bin)
    )
  )

plot_screen_dep <- ggplot(
  data_summary,
  aes(
    x = screening_status,
    y = proportion,
    fill = depression_status
  )
) +
  geom_col(width = 0.65, color = "white") +
  geom_text(
    aes(label = scales::percent(proportion, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    size = 4
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, .05))
  ) +
  scale_fill_manual(
    values = c(
      "No depressive symptoms" = "#4E79A7",
      "Postpartum depressive symptoms" = "#E15759"
    )
  ) +
  labs(
    title = "Postpartum Depressive Symptoms by Prenatal Mental Health Screening",
    subtitle = "PRAMS Phase 8",
    x = "Prenatal mental health screening",
    y = "Percent of respondents",
    fill = "Postpartum depressive symptoms"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

plot_screen_dep

ggsave(
  filename = here("results","figures","screening_vs_postpartum_depression.png"),
  plot = plot_screen_dep,
  width = 7,
  height = 5,
  dpi = 300
)
```

## Cross tabulation
```{r}
cross_tab <- analytic_dat |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(
    percent = round(n / sum(n) * 100, 1),
    screening_status = case_when(
      screened == 0 ~ "Not screened",
      screened == 1 ~ "Screened",
      TRUE ~ as.character(screened)
    ),
    depression_status = case_when(
      pp_depress_bin == 0 ~ "No depressive symptoms",
      pp_depress_bin == 1 ~ "Postpartum depressive symptoms",
      TRUE ~ as.character(pp_depress_bin)
    )
  ) |>
  select(screening_status, depression_status, n, percent)

cross_tab
```

## Interpretation
The exploratory analysis suggests that postpartum depressive symptoms were slightly less common among respondents who reported being screened during prenatal care. About 13.9% of screened respondents reported postpartum depressive symptoms compared with 15.2% among those who were not screened.

## Depression by prior depression during pregnancy
```{r}
dep_prior_tbl <- analytic_dat |>
  count(BPG_DEPRS8, pp_depress_bin) |>
  group_by(BPG_DEPRS8) |>
  mutate(pct = n / sum(n))

dep_prior_tbl
```

## Depression by education
```{r}
dep_edu_tbl <- analytic_dat |>
  count(MAT_ED, pp_depress_bin) |>
  group_by(MAT_ED) |>
  mutate(pct = n / sum(n))

dep_edu_tbl
```

## Proportion with postpartum depressive symptoms by screening
```{r}
plot_ppd_screen <- analytic_dat |>
  mutate(
    screening_status = case_when(
      screened == 0 ~ "Not screened",
      screened == 1 ~ "Screened",
      TRUE ~ as.character(screened)
    )
  ) |>
  ggplot(aes(x = screening_status,
             y = as.numeric(pp_depress_bin))) +
  stat_summary(fun = mean, geom = "bar", width = 0.65) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion with Postpartum Depressive Symptoms by Screening Status",
    x = "Prenatal mental health screening",
    y = "Percent with postpartum depressive symptoms"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold")
  )

plot_ppd_screen
```

## Save screening summary table
```{r}
screening_table <- analytic_dat |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(percent = round(n / sum(n) * 100, 1)) |>
  mutate(
    `Prenatal screening` = case_when(
      screened == 0 ~ "Not screened",
      screened == 1 ~ "Screened",
      TRUE ~ as.character(screened)
    ),
    `Postpartum depressive symptoms` = case_when(
      pp_depress_bin == 0 ~ "No",
      pp_depress_bin == 1 ~ "Yes",
      TRUE ~ as.character(pp_depress_bin)
    )
  ) |>
  select(
    `Prenatal screening`,
    `Postpartum depressive symptoms`,
    n,
    percent
  )

screening_table

write_csv(
  screening_table,
  here("results", "tables", "screening_summary.csv")
)
```