---
title: "Prenatal Mental Health Screening and Postpartum Depressive Symptoms"
author: "Alexandra Tejada-Strop"
bibliography: ../../assets/references/references.bib
csl: ../../assets/references/apa.csl

format:
  html:
    toc: true
    number-sections: true
    highlight-style: github

execute:
  echo: false
  warning: false
  message: false
  eval: true
---

```{r}
library(tidyverse)
library(here)
library(broom)
library(knitr)
library(scales)
library(survey)
library(kableExtra)

analytic_path <- here("data", "processed-data", "prams_analytic_localfix.rds")
stopifnot(file.exists(analytic_path))

analytic_dat <- readRDS(analytic_path)

n_total <- nrow(analytic_dat)

analytic_complete <- analytic_dat |>
  select(pp_depress_bin, screened, bpg_deprs8_bin, mat_ed) |>
  filter(bpg_deprs8_bin %in% c(0, 1)) |>
  drop_na()

n_analysis <- nrow(analytic_complete)

biv_tbl <- readRDS(here("results", "tables", "glm_bivariate_screened.rds"))
multi_tbl <- readRDS(here("results", "tables", "glm_multivariable_main.rds"))
weighted_tbl <- readRDS(here("results", "tables", "glm_weighted_main.rds"))
interaction_tbl <- readRDS(here("results", "tables", "glm_interaction_main.rds"))
stratified_tbl <- readRDS(here("results", "tables", "glm_stratified_main.rds"))
interaction_tbl <- readRDS(here("results", "tables", "glm_interaction_main.rds"))
stratified_tbl <- readRDS(here("results", "tables", "glm_stratified_main.rds"))

model_validation_path <- here("results", "tables", "model_comparison.rds")
if (file.exists(model_validation_path)) {
  model_validation_results <- readRDS(model_validation_path)
} else {
  model_validation_results <- tibble(
    model = c("Logistic regression", "Random forest"),
    .metric = c("accuracy", "roc_auc"),
    mean = c(NA_real_, NA_real_),
    std_err = c(NA_real_, NA_real_),
    n = c(NA_integer_, NA_integer_)
  )
}
```

Disclaimer: The opinions expressed in this article are the author's own and do not reflect their employer.

# Abstract
This study explores whether screening for mental health issues during pregnancy is linked to depression after giving birth, using data from the Pregnancy Risk Assessment Monitoring System (PRAMS) Phase 8. It mostly relies on women’s self-reports about their prenatal care and mental health afterward. The researchers used basic statistics and logistic regression to see if there is a connection between screening and postpartum depression, while considering other important factors. They found that women who were screened had lower chances of postpartum depression (OR = 0.76), which is an interesting finding that needs more research. Since the data is based on self-reports and observational, the results should be interpreted with caution.

# Introduction
Postpartum depression is an important health issue that affects a mother's well-being, family life, and the baby's health @bauman2020vitalsigns. Guidelines now recommend regular screening for depression and anxiety during pregnancy and after birth @acog2023screening; @uspstf2019perinatal.
Even with these guidelines, it is still unclear whether screening for mental health during pregnancy actually reduces the number of postpartum depression cases overall. Most evidence comes from clinical or intervention studies, with less data from general populations.
Mental health during pregnancy and after birth is a key part of a mother's overall health. Poor mental health during this time can cause problems for both mothers and babies, like difficulty caring for oneself and issues with early child development @acog2023screening. Because of this, healthcare providers are encouraged to check mental health regularly throughout pregnancy and after delivery.
Screening during pregnancy might help find women who are already showing symptoms or are at higher risk of postpartum depression. But we don’t have enough population-level data to fully understand how prenatal screening relates to postpartum depression, and learning more about this could improve clinical care and public health policies.
This study uses data from PRAMS, a surveillance system run by the CDC together with state health departments @cdc_prams_phase8_codebook; @cdc_prams_phase8_topics. PRAMS gathers survey responses from women who recently had a live birth and links these responses with birth certificate data. The survey asks about their experiences before, during, and after pregnancy, including prenatal care, health behaviors, mental health, and postpartum results. Forty-six states, the District of Columbia, New York City, the Northern Mariana Islands, and Puerto Rico participate, covering 81% of all births in the U.S. With thousands of responses, PRAMS provides valuable data for research and model building.

# Research Question
Do women who are asked about depression or anxiety during prenatal visits less often experience postpartum depression? The main thing we look at is whether women say they felt depressed or lost interest after giving birth, based on survey questions. The key factor is whether healthcare providers asked about mental health during pregnancy. Other factors include depression during pregnancy and the mother's education level.

# Methods
## Use of Computational Assistance
ChatGPT was used to help troubleshoot parts of the code. All analyses, decisions, and interpretations were reviewed and completed by the author.

## Analysis Approach
We first compared people who reported having prenatal mental health screening with those who did not. This helped us check the data quality, find missing details, and spot early patterns. Then, we used statistical methods to look at the link between prenatal screening and postpartum depression, considering important factors.

## Data Source
This data comes from PRAMS, a survey run by the Centers for Disease Control and Prevention and state health departments. PRAMS collects information people share about their experiences before, during, and shortly after pregnancy. Because PRAMS uses a detailed survey, the data can be analyzed with special variables included in the dataset. The main analyses here are unweighted, but a weighted analysis using PRAMS survey weights is also included to check how applicable the results are to the general population.

## Study Population
The full PRAMS analytic dataset included `r scales::comma(n_total)` respondents. After excluding observations with missing values for the primary exposure, outcome, and selected covariates, the final analytic sample included `r scales::comma(n_analysis)` respondents. Complete case analysis was used, which might cause bias if data are not missing completely at random. Respondents with missing data might differ in important ways from those with complete data, potentially affecting how well the results apply to other groups. Table 1 summarizes the characteristics of the study population.

```{r}
#| label: tbl-study-pop
#| tbl-cap: "Study population characteristics"

study_pop_tbl <- tibble(
  Characteristic = c(
    "Sample size",
    "  Total respondents",
    "  Respondents in analytic sample",
    "Prenatal screening",
    "  Screened during prenatal care",
    "  Not screened during prenatal care",
    "Postpartum depressive symptoms",
    "  No",
    "  Yes",
    "Depression during pregnancy",
    "  Yes",
    "  No",
    "Maternal education",
    "  Less than high school",
    "  High school graduate",
    "  Some college",
    "  College graduate",
    "  More than college"
  ),
  n = c(
    "",
    scales::comma(n_total),
    scales::comma(n_analysis),
    "",
    scales::comma(sum(analytic_complete$screened == 1)),
scales::comma(sum(analytic_complete$screened == 0)),
    "",
    scales::comma(sum(analytic_complete$pp_depress_bin == 0)),
    scales::comma(sum(analytic_complete$pp_depress_bin == 1)),
    "",
    scales::comma(sum(analytic_complete$bpg_deprs8_bin == 1)),
    scales::comma(sum(analytic_complete$bpg_deprs8_bin == 0)),
    "",
    scales::comma(sum(analytic_complete$mat_ed == 1)),
    scales::comma(sum(analytic_complete$mat_ed == 2)),
    scales::comma(sum(analytic_complete$mat_ed == 3)),
    scales::comma(sum(analytic_complete$mat_ed == 4)),
    scales::comma(sum(analytic_complete$mat_ed == 5))
  ),
  Percent = c(
    "",
    "100.0%",
    paste0(round(100 * n_analysis / n_total, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$screened == 1) / n_analysis, 1), "%"),
paste0(round(100 * sum(analytic_complete$screened == 0) / n_analysis, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$pp_depress_bin == 0) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$pp_depress_bin == 1) / n_analysis, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$bpg_deprs8_bin == 1) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$bpg_deprs8_bin == 0) / n_analysis, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$mat_ed == 1) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 2) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 3) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 4) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 5) / n_analysis, 1), "%")
  )
)

knitr::kable(study_pop_tbl, align = c("l", "r", "r"), booktabs = TRUE) |>
  kableExtra::kable_styling(full_width = FALSE, position = "left", font_size = 11)

```

## Measures
The main exposure was whether a healthcare provider talked about depression or anxiety during prenatal visits, based on a survey question. This was recorded as yes or no.
The main outcome was postpartum depressive symptoms, measured using survey questions about feelings of depression and loss of interest after childbirth. These questions have been used before in research on postpartum depression @bauman2020vitalsigns; @farr2016provider.
Key factors were depression during pregnancy and the mother's education level. Depression during pregnancy was included because women who have had depression before are more likely to have depressive symptoms after giving birth. The mother's education was included because it can affect access to healthcare and mental health.

## Statistical Approach
Exploratory analysis included basic statistics, cross-tabulation, and bar charts to compare postpartum depression symptoms based on screening status. All of this was done using R.

A multivariable logistic regression was used to see if postpartum depression symptoms were related to prenatal mental health screening. We also considered depression during pregnancy and maternal education as additional factors. The results are shown as odds ratios with 95% confidence intervals.

A random forest model was also run as a sensitivity check to assess whether a more flexible approach improved predictions over logistic regression. Full model validation results are reported in the Supplemental Materials.

To account for the PRAMS survey design, a weighted logistic regression was run using the survey package in R, incorporating survey weights, cluster IDs, and strata variables to produce population-representative estimates.

To see if the effect of screening was different for women who did or did not have depression during pregnancy, we added an interaction term between screening and depression. Lastly, separate logistic regressions were run for women with and without depression during pregnancy to examine the effect of screening within each group.

# Results
The analytic sample included `r scales::comma(n_analysis)` respondents. Overall, `r round(100 * sum(analytic_complete$screened == 1) / n_analysis, 1)`% reported being screened for depression during prenatal care, and `r round(100 * sum(analytic_complete$pp_depress_bin == 1) / n_analysis, 1)`% reported postpartum depressive symptoms.

## Exploratory Analysis
Postpartum depression was a bit more common among women who weren’t screened, at 15.1%, compared to 13.9% for those who were. Women who had depression during pregnancy also showed higher rates. There were clear differences across education levels. Table 2 shows that, before adjusting for other factors, women who were screened already had lower chances of postpartum depression than those who were not.

```{r}
#| label: tbl-bivariate
#| tbl-cap: "Unadjusted odds ratio for prenatal mental health screening and postpartum depressive symptoms"

biv_clean <- biv_tbl |>
  filter(term == "screened1") |>
  transmute(
    Predictor = "Prenatal mental health screening",
    `Odds ratio` = sprintf("%.2f", estimate),
    `95% CI` = paste0("(", sprintf("%.2f", conf.low), ", ", sprintf("%.2f", conf.high), ")"),
    `P value` = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  )

knitr::kable(
  biv_clean,
  col.names = c("Predictor", "Odds ratio", "95% CI", "P value"),
  align = c("l", "c", "c", "c"),
  booktabs = TRUE
) |>
  kableExtra::kable_styling(full_width = FALSE, position = "left") |>
  kableExtra::footnote(
    general = "Unadjusted model. Reference group: not screened during prenatal care.",
    general_title = "Note:",
    footnote_as_chunk = TRUE
  )
```

Before considering other factors, the respondents who were screened already had lower chances of postpartum depression compared to those who were not screened. Table 3 presents the adjusted odds ratio for all predictors in the model.

```{r}
#| label: tbl-multivariable
#| tbl-cap: "Adjusted odds ratios for factors associated with postpartum depressive symptoms"
multi_clean <- multi_tbl |>
  filter(term != "(Intercept)") |>
  mutate(
    Predictor = case_when(
      term == "screened1" ~ "Prenatal mental health screening",
      term == "bpg_deprs8_bin1" ~ "Depression during pregnancy",
      term == "mat_ed2" ~ "Education: High school graduate",
      term == "mat_ed3" ~ "Education: Some college",
      term == "mat_ed4" ~ "Education: College graduate",
      term == "mat_ed5" ~ "Education: More than college",
      TRUE ~ term
    ),
    OR = estimate,
    CI_low = conf.low,
    CI_high = conf.high
  ) |>
  transmute(
    Predictor,
    `Odds ratio` = sprintf("%.2f", OR),
    `95% CI` = paste0("(", sprintf("%.2f", CI_low), ", ", sprintf("%.2f", CI_high), ")"),
    `P value` = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  )

knitr::kable(
  multi_clean,
  col.names = c("Predictor", "Odds ratio", "95% CI", "P value"),
  align = c("l", "c", "c", "c"),
  booktabs = TRUE
) |>
  kableExtra::kable_styling(full_width = FALSE, position = "left") |>
  kableExtra::footnote(
    general = "Reference groups: not screened during prenatal care, no depression during pregnancy, less than high school education.",
    general_title = "Note:",
    footnote_as_chunk = TRUE
  )
```

```{r}
#| label: fig-screening
#| fig-cap: "Postpartum depressive symptoms by prenatal mental health screening"

fig_data <- analytic_complete |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    Screening = ifelse(screened == 0, "Not screened", "Screened"),
    Outcome = ifelse(pp_depress_bin == 0, "No depressive symptoms", "Depressive symptoms")
  )

ggplot(fig_data, aes(x = Screening, y = percent, fill = Outcome)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Depressive symptoms" = "#D55E00",
      "No depressive symptoms" = "#0072B2"
    )
  ) +
  labs(
    x = "Prenatal mental health screening",
    y = "Percent of respondents",
    fill = "Outcome"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )
```

Figure 1 shows the distribution of postpartum depressive symptoms by prenatal screening status.

```{r}
#| label: fig-depression-pregnancy
#| fig-cap: "Postpartum depressive symptoms by depression during pregnancy"

fig_dep_preg <- analytic_complete |>
  filter(bpg_deprs8_bin %in% c(0, 1), pp_depress_bin %in% c(0, 1)) |>
  count(bpg_deprs8_bin, pp_depress_bin) |>
  group_by(bpg_deprs8_bin) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    Depression_pregnancy = ifelse(bpg_deprs8_bin == 1, "Yes", "No"),
    Outcome = ifelse(pp_depress_bin == 0, "No depressive symptoms", "Depressive symptoms")
  )

ggplot(fig_dep_preg, aes(x = Depression_pregnancy, y = percent, fill = Outcome)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Depressive symptoms" = "#D55E00",
      "No depressive symptoms" = "#0072B2"
    )
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Depression during pregnancy",
    y = "Percent of respondents",
    fill = "Outcome"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

dir.create(here("results", "figures"), showWarnings = FALSE, recursive = TRUE)

ggsave(
  filename = here("results", "figures", "depression_pregnancy_vs_postpartum.png"),
  plot = last_plot(),
  width = 8,
  height = 5,
  dpi = 300
)
```

Women who experience depression during pregnancy are more likely to have postpartum depression. About 31 out of 100 women with depression during pregnancy also had postpartum depression, compared to about 11 out of 100 women who did not experience depression.

```{r}
#| label: tbl-screening
#| tbl-cap: "Distribution of postpartum depressive symptoms by prenatal mental health screening"

screening_table <- analytic_complete |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(
    percent = round(100 * n / sum(n), 1),
    Screening = ifelse(screened == 0, "Not screened", "Screened"),
    Outcome = ifelse(pp_depress_bin == 0, "No depressive symptoms", "Depressive symptoms")
  ) |>
  ungroup() |>
  mutate(value = paste0(scales::comma(n), " (", percent, "%)")) |>
  select(Screening, Outcome, value) |>
  pivot_wider(names_from = Outcome, values_from = value) |>
  rename(`Screening status` = Screening) |>
  select(`Screening status`, `Depressive symptoms`, `No depressive symptoms`)

knitr::kable(screening_table, align = c("l", "c", "c"))
```

Table 4 shows that 15.1% of respondents who were not screened reported postpartum depression symptoms, compared to 13.9% of those who were screened. This aligns with the results from the adjusted model.


## Weighted Analysis
To see if the results apply to the wider group of new mothers, the main model was re-run using survey weights from PRAMS. The results were similar to the original analysis without weights. Women who were screened had lower chances of postpartum depression (OR = 0.79, 95% CI: 0.75 to 0.82, p < 0.001), and depression during pregnancy was a strong predictor (OR = 3.46, 95% CI: 3.30 to 3.62, p < 0.001). The similarity between weighted and unweighted results suggests the original findings are likely representative of the larger population.

```{r}
#| label: tbl-weighted-comparison
#| tbl-cap: "Comparison of unweighted and weighted odds ratios for main predictors"

comparison_tbl <- tibble(
  Predictor = c(
    "Prenatal mental health screening",
    "Depression during pregnancy",
    "Education: High school graduate",
    "Education: Some college",
    "Education: College graduate",
    "Education: More than college"
  ),
  `Unweighted OR` = c("0.76 (0.74, 0.78)", "3.38 (3.29, 3.47)", 
                       "1.36 (1.25, 1.48)", "1.29 (1.19, 1.39)",
                       "1.09 (1.01, 1.18)", "0.69 (0.64, 0.75)"),
  `Weighted OR` = c("0.79 (0.75, 0.82)", "3.46 (3.30, 3.62)",
                     "1.41 (1.23, 1.62)", "1.29 (1.13, 1.48)",
                     "1.12 (0.99, 1.28)", "0.70 (0.61, 0.80)")
)

knitr::kable(
  comparison_tbl,
  col.names = c("Predictor", "Unweighted OR (95% CI)", "Weighted OR (95% CI)"),
  align = c("l", "c", "c")
) |>
  kableExtra::kable_styling(full_width = FALSE, position = "left")

```


## Interaction Analysis
To see if the effect of prenatal screening was different for women with a history of depression during pregnancy, an interaction term was added to the model. The results showed that there was no significant difference between the groups (OR = 1.02, 95% CI: 0.95 to 1.10, p = 0.58). This suggests that the protective effect of screening is similar for women regardless of their depression history during pregnancy.

## Stratified Analysis
Logistic regression analysis showed that women who were screened during pregnancy had lower chances of experiencing postpartum depression, whether or not they had depression during pregnancy. For women without depression, the odds were about 24% lower, and for those with depression, about 23% lower. These findings suggest that screening during pregnancy can be helpful for all women, no matter their mental health history.

```{r}
#| label: fig-stratified
#| fig-cap: "Postpartum depressive symptoms by screening status and depression during pregnancy"

strat_plot_data <- analytic_complete |>
  mutate(
    depression_group = ifelse(bpg_deprs8_bin == 1, 
                              "Depression during pregnancy", 
                              "No depression during pregnancy"),
    screening_status = ifelse(screened == 1, "Screened", "Not screened")
  ) |>
  group_by(depression_group, screening_status) |>
  summarise(
    pct_ppd = mean(pp_depress_bin == 1) * 100,
    .groups = "drop"
  )

ggplot(strat_plot_data, 
       aes(x = depression_group, y = pct_ppd, fill = screening_status)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(
    aes(label = paste0(round(pct_ppd, 1), "%")),
    position = position_dodge(width = 0.6),
    vjust = -0.5,
    size = 3.5
  ) +
  scale_fill_manual(
    values = c("Not screened" = "#D55E00", "Screened" = "#0072B2")
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    limits = c(0, 40)
  ) +
  labs(
    x = "Depression during pregnancy",
    y = "Percent with postpartum depressive symptoms",
    fill = "Screening status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )
```


# Discussion
This study looked at whether screening for mental health issues during pregnancy was linked to a lower risk of depression after giving birth, based on PRAMS data. Women who were screened were less likely to experience postpartum depression, even when accounting for past depression and their education. However, since this was an observational study, we can't be certain that the screening directly caused the lower depression rates. One reason might be that screening helps women understand mental health concerns better and encourages them to seek help early, connecting them with support before delivery. Another reason could be that screening identifies women who were already more at risk for depression, which might show up later regardless of what actions were taken.
These findings are broadly consistent with prior work showing that postpartum depressive symptoms are common and that provider screening practices vary @bauman2020vitalsigns; @farr2016provider. Clinical guidelines recommend routine screening during pregnancy and postpartum @acog2023screening, and these results add to the population-level evidence on how screening relates to reported symptoms.

The biggest thing we found was how strongly depression during pregnancy predicted postpartum depression symptoms. Women who experienced depression while pregnant were more than three times as likely to have postpartum symptoms (OR = 3.38, weighted OR = 3.46). This shows how important it is to consider mental health history when looking at postpartum outcomes. Interestingly, the analysis showed that the protective effect of screening was similar for women with and without depression during pregnancy (OR = 1.02, p = 0.58). The separate analysis confirmed this, with similar screening ORs in both groups (0.76 and 0.77). These results support universal mental health screening during pregnancy rather than only focusing on high-risk women.
It is important to remember what this study cannot tell us. Since the data come from just one point in time and are based on women’s own reports, we cannot say if screening directly led to better results. Women who were screened might have had better access to healthcare or more support from providers, which could also affect postpartum mental health.
**Strengths:** PRAMS is a large dataset that includes data from multiple states, providing enough information for detailed analysis and making the results more relevant across different areas than studies from just one place.
**Limitations:** The data come from self-reports, which can be affected by memory errors. The measure for postpartum depression symptoms is not a hospital diagnosis. Since the data are from one point in time, it is hard to know what causes what. Also, there are factors we did not measure, like whether someone has insurance, their social support, or how healthcare providers deliver care, which could influence the results.
In future research, including details like insurance status, access to mental health services, and income could be helpful. This study also used weighted analysis with PRAMS survey weights, and the results matched the unweighted estimates, which supports that the findings can be applied more broadly.

# References



