---
title: "Prenatal Mental Health Screening and Postpartum Depressive Symptoms"
author: "Alexandra Tejada-Strop"
bibliography: references.bib
csl: apa.csl

format:
  html:
    toc: true
    number-sections: true
    highlight-style: github

execute:
  echo: false
  warning: false
  message: false
  eval: true
---

```{r}
library(tidyverse)
library(here)
library(broom)
library(knitr)
library(scales)
library(survey)
library(kableExtra)

analytic_path <- here("data", "processed-data", "prams_analytic.rds")
stopifnot(file.exists(analytic_path))

analytic_dat <- readRDS(analytic_path)

dir.create(here("class-project", "figures"), showWarnings = FALSE, recursive = TRUE)

n_total <- nrow(analytic_dat)

analytic_complete <- analytic_dat |>
  select(pp_depress_bin, screened, bpg_deprs8, mat_ed) |>
  drop_na()

n_analysis <- nrow(analytic_complete)

biv_tbl <- readRDS(here("results", "tables", "glm_bivariate_screened.rds"))
multi_tbl <- readRDS(here("results", "tables", "glm_multivariable_main.rds"))

model_validation_path <- here("results", "tables", "model_comparison.rds")
if (file.exists(model_validation_path)) {
  model_validation_results <- readRDS(model_validation_path)
} else {
  model_validation_results <- tibble(
    model = c("Logistic regression", "Random forest"),
    .metric = c("accuracy", "roc_auc"),
    mean = c(NA_real_, NA_real_),
    std_err = c(NA_real_, NA_real_),
    n = c(NA_integer_, NA_integer_)
  )
}
```

Disclaimer: The opinions expressed in this article are the author's own and do not reflect their employer.

# Abstract
This study examines whether prenatal mental health screening is associated with postpartum depressive symptoms using data from the Pregnancy Risk Assessment Monitoring System, or PRAMS, Phase 8. The analysis focuses on self reported prenatal care experiences and postpartum mental health outcomes.

Descriptive statistics and multivariable logistic regression were used to evaluate whether screening is associated with differences in reported depressive symptoms while adjusting for key covariates. Prenatal mental health screening was associated with lower odds of postpartum depressive symptoms, although the magnitude of the association was modest. Findings should be interpreted cautiously because the data are observational and self reported.

# Introduction
Postpartum depression is a significant public health concern that affects maternal well being, family functioning, and infant outcomes @bauman2020vitalsigns. Clinical guidelines increasingly recommend routine screening for depression and anxiety during pregnancy and the postpartum period @acog2023screening; @uspstf2019perinatal.

Despite these recommendations, it remains unclear whether prenatal mental health screening is associated with postpartum depressive symptoms in real world population settings. Much of the existing evidence comes from clinical or intervention based contexts, with less emphasis on population based surveillance data.

Using PRAMS Phase 8 data, this study evaluates whether individuals who report being asked about depression or anxiety during prenatal care are less likely to report postpartum depressive symptoms.

Prenatal and postpartum mental health conditions are increasingly recognized as critical components of maternal health. Poor maternal mental health has been associated with adverse outcomes for both mothers and infants, including reduced maternal functioning and challenges in early child development @acog2023screening. As a result, healthcare providers are encouraged to assess mental health concerns throughout pregnancy and after delivery.

Routine screening for depression during pregnancy may provide opportunities for earlier identification of individuals experiencing symptoms or those at elevated risk of developing postpartum depression. However, population level evidence examining how prenatal screening relates to postpartum depressive symptoms remains limited. Understanding this relationship may help inform both clinical practice and maternal health policy.

The data for this study come from the Pregnancy Risk Assessment Monitoring System, a population based surveillance system conducted by the Centers for Disease Control and Prevention in collaboration with state health departments @cdc_prams_phase8_codebook; @cdc_prams_phase8_topics. PRAMS collects self reported survey data from individuals who recently had a live birth and links these responses to selected birth certificate variables.

PRAMS is designed to monitor maternal experiences and behaviors before, during, and shortly after pregnancy. The survey includes questions related to prenatal care, maternal health conditions, health behaviors, mental health, and postpartum experiences. These data allow researchers to examine population level patterns related to maternal and infant health.

PRAMS Phase 8 includes several hundred variables capturing maternal demographics, pre pregnancy health conditions, prenatal care experiences, health behaviors, insurance coverage, mental health indicators, and postpartum outcomes. The dataset contains a mix of categorical and numeric variables and reflects the real world complexity of survey data, including skip patterns and missingness. Across participating states, PRAMS Phase 8 includes tens of thousands of respondents, providing sufficient sample size and variable richness to support exploratory analysis and regression modeling.

# Research Question
The primary research question for this study is: Are individuals who report being asked about depression or anxiety during prenatal care less likely to report postpartum depressive symptoms?

The primary outcome of interest is self reported postpartum depressive symptoms, measured using PRAMS survey items assessing feelings of depression and loss of interest since birth. The primary predictor of interest is prenatal mental health screening, defined by whether a healthcare provider asked about depression or anxiety during prenatal visits. Additional covariates considered for analysis include depression during pregnancy and maternal education.

# Methods
## Use of Computational Assistance
Artificial intelligence tool ChatGPT was used to assist with code troubleshooting. All code and analytical decisions were reviewed, modified as needed, and implemented by the author. The author takes full responsibility for the accuracy, interpretation, and reproducibility of all analyses and results.

# Analysis Approach
The analysis plan included descriptive summaries of key variables and exploratory comparisons between individuals who did and did not report prenatal mental health screening. These exploratory analyses were used to assess data quality, missingness, and preliminary patterns in the data.

Regression based and model validation approaches were then used to examine associations between prenatal screening and postpartum mental health outcomes while accounting for relevant covariates.

## Data Source
Data were obtained from PRAMS, a population based surveillance system conducted by the Centers for Disease Control and Prevention in collaboration with state health departments. PRAMS collects self reported information on maternal experiences before, during, and shortly after pregnancy.

Because PRAMS uses a complex survey design, weighted analyses can be conducted using the survey design variables provided in the dataset. However, the results presented here focus on descriptive summaries and logistic regression models based on the analytic dataset used for this course project.

## Study Population
The full PRAMS analytic dataset included `r scales::comma(n_total)` respondents. After excluding observations with missing values for the primary exposure, outcome, and selected covariates used in the main model, the final analytic sample included `r scales::comma(n_analysis)` respondents.

```{r}
#| label: tbl-study-pop
#| tbl-cap: "Study population characteristics"

study_pop_tbl <- tibble(
  Characteristic = c(
    "Sample size",
    "  Total respondents",
    "  Respondents in analytic sample",
    "Prenatal screening",
    "  Screened during prenatal care",
    "  Not screened during prenatal care",
    "Postpartum depressive symptoms",
    "  Yes",
    "  No",
    "Depression during pregnancy",
    "  No",
    "  Yes",
    "Maternal education",
    "  Less than high school",
    "  High school graduate",
    "  Some college",
    "  College graduate",
    "  More than college"
  ),
  n = c(
    "",
    scales::comma(n_total),
    scales::comma(n_analysis),
    "",
    scales::comma(sum(analytic_complete$screened == 1)),
    scales::comma(sum(analytic_complete$screened == 0)),
    "",
    scales::comma(sum(analytic_complete$pp_depress_bin == 1)),
    scales::comma(sum(analytic_complete$pp_depress_bin == 0)),
    "",
    scales::comma(sum(analytic_complete$bpg_deprs8 == 1)),
    scales::comma(sum(analytic_complete$bpg_deprs8 == 2)),
    "",
    scales::comma(sum(analytic_complete$mat_ed == 1)),
    scales::comma(sum(analytic_complete$mat_ed == 2)),
    scales::comma(sum(analytic_complete$mat_ed == 3)),
    scales::comma(sum(analytic_complete$mat_ed == 4)),
    scales::comma(sum(analytic_complete$mat_ed == 5))
  ),
  Percent = c(
    "",
    "100.0%",
    paste0(round(100 * n_analysis / n_total, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$screened == 1) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$screened == 0) / n_analysis, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$pp_depress_bin == 1) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$pp_depress_bin == 0) / n_analysis, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$bpg_deprs8 == 1) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$bpg_deprs8 == 2) / n_analysis, 1), "%"),
    "",
    paste0(round(100 * sum(analytic_complete$mat_ed == 1) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 2) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 3) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 4) / n_analysis, 1), "%"),
    paste0(round(100 * sum(analytic_complete$mat_ed == 5) / n_analysis, 1), "%")
  )
)

knitr::kable(study_pop_tbl, align = c("l", "r", "r"), booktabs = TRUE) |>
  kableExtra::kable_styling(full_width = FALSE, position = "left", font_size = 11)
```

## Measures
The primary exposure in this study was self reported prenatal mental health screening, measured using a PRAMS survey item indicating whether a healthcare provider asked about depression or anxiety during prenatal care. This variable was coded as a binary indicator representing whether respondents reported being screened during prenatal visits.

The primary outcome was postpartum depressive symptoms. PRAMS includes survey items assessing whether respondents experienced feelings consistent with depression or loss of interest following childbirth. These items have been used in previous research examining postpartum depressive symptoms using PRAMS data @bauman2020vitalsigns; @farr2016provider.

Key covariates included depression during pregnancy and maternal education. Depression during pregnancy was included as an important potential confounder, as individuals with a history of depression may be more likely to experience postpartum depressive symptoms. Maternal education was included as a sociodemographic variable that may influence both healthcare experiences and mental health outcomes.

## Statistical Approach
Exploratory analyses included descriptive statistics, cross tabulations, and graphical comparisons of postpartum depressive symptoms by screening status. All analyses were conducted in R.

Multivariable logistic regression models were specified with postpartum depressive symptoms as the binary outcome. Prenatal mental health screening was included as the primary exposure. Covariates included depression during pregnancy and maternal education. Results are presented as odds ratios with corresponding 95 percent confidence intervals.

\newpage

# Results
The analytic sample included `r scales::comma(n_analysis)` respondents after excluding observations with missing values for the primary exposure, outcome, and selected covariates. Overall, `r round(100 * sum(analytic_complete$screened == 1) / n_analysis, 1)` percent of respondents reported being screened for depression during prenatal care, and `r round(100 * sum(analytic_complete$pp_depress_bin == 1) / n_analysis, 1)` percent reported postpartum depressive symptoms.

## Exploratory Analysis
We first examined the distribution of prenatal mental health screening and postpartum depressive symptoms in the analytic sample. We also explored how postpartum depressive symptoms varied by screening status, depression during pregnancy, and maternal education.

Descriptive results suggested that postpartum depressive symptoms were more common among respondents who did not report receiving prenatal mental health screening. Higher rates of postpartum depressive symptoms were also observed among respondents who reported depression during pregnancy. Differences across maternal education groups were also examined to assess potential patterns in the data.

These exploratory analyses helped identify initial patterns in the dataset and informed the selection of variables for adjusted modeling.

```{r}
#| label: tbl-multivariable
#| tbl-cap: "Adjusted odds ratios for postpartum depressive symptoms"

multi_clean <- multi_tbl |>
  filter(term != "(Intercept)") |>
  mutate(
    Predictor = case_when(
      term == "screened1" ~ "Screened during prenatal care",
      term == "bpg_deprs82" ~ "Depression during pregnancy",
      term == "mat_ed2" ~ "High school graduate",
      term == "mat_ed3" ~ "Some college",
      term == "mat_ed4" ~ "College graduate",
      term == "mat_ed5" ~ "More than college",
      TRUE ~ term
    ),
    OR = estimate,
    CI_low = conf.low,
    CI_high = conf.high
  ) |>
  transmute(
    Predictor,
    `Odds ratio` = sprintf("%.2f", OR),
    `95% CI` = paste0("(", sprintf("%.2f", CI_low), ", ", sprintf("%.2f", CI_high), ")"),
    `P value` = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  )

knitr::kable(
  multi_clean,
  col.names = c("Predictor", "Odds ratio", "95% CI", "P value"),
  align = c("l", "c", "c", "c"),
  booktabs = TRUE
) |>
  kableExtra::kable_styling(full_width = FALSE, position = "left")
```


```{r}
#| label: fig-screening
#| fig-cap: "Postpartum depressive symptoms by prenatal mental health screening"

fig_data <- analytic_complete |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    Screening = ifelse(screened == 1, "Screened", "Not screened"),
    Outcome = ifelse(pp_depress_bin == 1, "Depressive symptoms", "No depressive symptoms")
  )

ggplot(fig_data, aes(x = Screening, y = percent, fill = Outcome)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Depressive symptoms" = "#D55E00",
      "No depressive symptoms" = "#0072B2"
    )
  ) +
  labs(
    x = "Prenatal mental health screening",
    y = "Percent of respondents",
    fill = "Outcome"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )
```


```{r}
#| label: fig-depression-pregnancy
#| fig-cap: "Postpartum depressive symptoms by depression during pregnancy"

fig_dep_preg <- analytic_complete |>
  filter(bpg_deprs8 %in% c(1, 2), pp_depress_bin %in% c(0, 1)) |>
  count(bpg_deprs8, pp_depress_bin) |>
  group_by(bpg_deprs8) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    Depression_pregnancy = ifelse(bpg_deprs8 == 2, "Yes", "No"),
    Outcome = ifelse(pp_depress_bin == 1, "Depressive symptoms", "No depressive symptoms")
  )

ggplot(fig_dep_preg, aes(x = Depression_pregnancy, y = percent, fill = Outcome)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Depressive symptoms" = "#D55E00",
      "No depressive symptoms" = "#0072B2"
    )
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Depression during pregnancy",
    y = "Percent of respondents",
    fill = "Outcome"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

dir.create(here("class-project", "figures"), showWarnings = FALSE, recursive = TRUE)  

ggsave(
  filename = here("class-project", "figures", "depression_pregnancy_vs_postpartum.png"),
  plot = last_plot(),
  width = 8,
  height = 5,
  dpi = 300
)
```


In exploratory analyses, postpartum depressive symptoms were more common among respondents who did not report receiving prenatal depression screening compared with those who were screened. Depression during pregnancy was strongly associated with postpartum depressive symptoms, suggesting that mental health history is an important confounder to consider in adjusted analyses. Differences in postpartum depressive symptoms were also observed across education levels.

```{r}
#| label: tbl-screening
#| tbl-cap: "Distribution of postpartum depressive symptoms by prenatal mental health screening"

screening_table <- analytic_complete |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(
    percent = round(100 * n / sum(n), 1),
    Screening = ifelse(screened == 1, "Screened", "Not screened"),
    Outcome = ifelse(pp_depress_bin == 1, "Depressive symptoms", "No depressive symptoms")
  ) |>
  ungroup() |>
  mutate(value = paste0(scales::comma(n), " (", percent, "%)")) |>
  select(Screening, Outcome, value) |>
  pivot_wider(names_from = Outcome, values_from = value) |>
  rename(`Screening status` = Screening)

knitr::kable(screening_table, align = c("l", "c", "c"))
```

The table on postpartum depressive symptoms by screening status shows that 15.2 percent of respondents who were not screened during prenatal care reported postpartum depressive symptoms, compared with 13.9 percent among those who were screened.

```{r}
#| label: tbl-model-validation
#| tbl-cap: "Model validation results comparing logistic regression and random forest"

model_validation_tbl <- model_validation_results |>
  filter(.metric %in% c("accuracy", "roc_auc")) |>
  mutate(
    metric_label = case_when(
      .metric == "accuracy" ~ "Accuracy",
      .metric == "roc_auc" ~ "ROC AUC",
      TRUE ~ .metric
    ),
    mean = round(mean, 3)
  ) |>
  select(model, metric_label, mean) |>
  pivot_wider(names_from = metric_label, values_from = mean) |>
  rename(Model = model)

knitr::kable(model_validation_tbl, align = c("l", "c", "c"))
```


Reference groups were not screened during prenatal care, no depression during pregnancy, and less than high school education.

In adjusted analyses, respondents who reported prenatal mental health screening had lower odds of postpartum depressive symptoms compared with those who were not screened. In the multivariable model, prenatal screening was associated with lower odds of postpartum depressive symptoms after adjustment for depression during pregnancy and maternal education.

Model validation results indicated that logistic regression and random forest models performed similarly. Because the random forest did not meaningfully improve predictive performance, logistic regression was retained as the preferred model due to its interpretability.

# Discussion
This study evaluated the association between prenatal mental health screening and postpartum depressive symptoms using population based PRAMS data. Respondents who reported prenatal mental health screening had lower odds of postpartum depressive symptoms than those who were not screened, even after adjustment for depression during pregnancy and maternal education.

Although PRAMS includes survey weights and design variables, the analyses presented here are unweighted and intended as an exploratory course project focused on reproducible modeling workflows.

These findings are generally consistent with existing research showing that postpartum depressive symptoms remain a significant public health concern in the United States. Previous analyses using PRAMS data have documented that a substantial proportion of women report postpartum depressive symptoms and that provider discussions about perinatal depression vary across populations and healthcare settings @bauman2020vitalsigns; @farr2016provider. Clinical guidelines from organizations such as the American College of Obstetricians and Gynecologists recommend routine screening for depression during pregnancy and the postpartum period to support earlier identification and intervention @acog2023screening.

The exploratory results from this study suggest that prenatal screening may be associated with differences in postpartum mental health outcomes at the population level. However, these results should be interpreted cautiously. Because the data are observational and based on self reported survey responses, the analyses cannot determine whether prenatal screening itself reduces postpartum depressive symptoms. It is possible that individuals who receive more comprehensive prenatal care are also more likely to receive screening and additional support services that influence postpartum mental health outcomes.

Depression during pregnancy emerged as one of the strongest predictors of postpartum depressive symptoms in the adjusted analyses. This finding aligns with previous research showing that mental health history is an important risk factor for postpartum depression. Accounting for prior depression is therefore essential when examining associations between prenatal care experiences and postpartum mental health outcomes.

Several strengths of this study should be noted. PRAMS provides a large, population based dataset that captures maternal experiences across multiple states. The survey includes detailed information on prenatal care experiences, health behaviors, and postpartum outcomes, allowing researchers to examine mental health patterns at a population level. The large sample size also supports exploratory regression analyses and the evaluation of multiple potential confounders.

This study also has several limitations. First, PRAMS data are self reported and may be subject to recall bias or reporting bias. Second, the postpartum depressive symptom measure used in PRAMS is not a clinical diagnosis of depression but rather a survey based indicator of depressive symptoms. Third, the cross sectional nature of the survey limits the ability to determine temporal relationships or causal effects. Finally, unmeasured factors such as healthcare access, provider practices, or social support may influence both screening and postpartum mental health outcomes.

Despite these limitations, the findings highlight the potential value of examining prenatal mental health screening practices using population based surveillance data. Understanding how screening practices relate to maternal mental health outcomes may help inform future research and public health efforts aimed at improving maternal mental health.

Future analyses could expand on this work by incorporating additional covariates such as insurance status, socioeconomic indicators, and access to mental health services. More advanced modeling approaches could also be used to explore predictors of postpartum depressive symptoms and identify subgroups that may benefit most from screening and early intervention.

Overall, this study demonstrates how large surveillance datasets such as PRAMS can be used to explore maternal mental health patterns and generate hypotheses for future research. Continued attention to perinatal mental health remains important for improving maternal and infant health outcomes.

\newpage

# References