---
title: "Prenatal Mental Health Screening and Postpartum Depressive Symptoms"
author: "Alexandra Tejada-Strop"
bibliography: prenatal_refs.bib
csl: apa.csl

format:
  html:
    toc: true
    number-sections: true
    highlight-style: github

execute:
  echo: false
  warning: false
  message: false
  eval: true
---

```{r}
library(tidyverse)
library(here)
library(broom)
library(knitr)
library(scales)
library(survey)

# Load analytic dataset
analytic_path <- here("data", "processed-data", "prams_analytic.rds")
analytic_dat <- readRDS(analytic_path)

# Sample sizes
n_total <- nrow(analytic_dat)

n_analysis <- analytic_dat |>
  select(pp_depress_bin, screened, BPG_DEPRS8, MAT_ED) |>
  drop_na() |>
  nrow()

# Load saved model results
biv_tbl <- readRDS(here("results", "tables", "glm_bivariate_screened.rds"))
multi_tbl <- readRDS(here("results", "tables", "glm_multivariable_main.rds"))

# Path to saved figure
fig_path <- here("results", "figures", "screening_vs_postpartum_depression.png")
```


$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer.


# Abstract
This study examines the association between prenatal mental health screening and postpartum depressive symptoms using data from the Pregnancy Risk Assessment Monitoring System (PRAMS) Phase 8. The analysis focuses on self-reported prenatal care experiences and postpartum mental health outcomes. Descriptive statistics and multivariable logistic regression were used to evaluate whether screening is associated with differences in reported depressive symptoms while adjusting for key covariates. Prenatal mental health screening was associated with lower odds of postpartum depressive symptoms, although the magnitude of the association was modest. Findings should be interpreted cautiously due to the observational and self-reported nature of the data.


# Introduction

Postpartum depression is a significant public health concern that affects maternal well-being, family functioning, and infant outcomes@bauman2020vitalsigns. Clinical guidelines increasingly recommend routine screening for depression and anxiety during pregnancy and the postpartum period @acog2023screening @uspstf2019perinatal.

Despite these recommendations, it remains unclear whether prenatal mental health screening is associated with postpartum depressive symptoms in real-world population settings. Most existing evidence focuses on clinical or intervention-based settings, with less emphasis on population-based surveillance data.

Using PRAMS Phase 8 data, this study evaluates whether individuals who report being asked about depression or anxiety during prenatal care are less likely to report postpartum depressive symptoms.


## General Background Information

Prenatal and postpartum mental health conditions are increasingly recognized as critical components of maternal health. Poor maternal mental health has been associated with adverse outcomes for both mothers and infants, including reduced maternal functioning and challenges in early child development @acog2023screening. As a result, healthcare providers are encouraged to assess mental health concerns throughout pregnancy and after delivery.

Routine screening for depression during pregnancy may provide opportunities for earlier identification of individuals experiencing symptoms or those at elevated risk of developing postpartum depression. However, population-level evidence examining how prenatal screening relates to postpartum depressive symptoms remains limited. Understanding this relationship may help inform both clinical practice and maternal health policy.

## Description of data

The data for this study come from the Pregnancy Risk Assessment Monitoring System (PRAMS) Phase 8, a population-based surveillance system conducted by the Centers for Disease Control and Prevention in collaboration with state health departments @cdc_prams_phase8_codebook @cdc_prams_phase8_topics. PRAMS collects self-reported survey data from individuals who recently had a live birth and links these responses to selected birth certificate variables.

PRAMS is designed to monitor maternal experiences and behaviors before, during, and shortly after pregnancy. The survey includes questions related to prenatal care, maternal health conditions, health behaviors, mental health, and postpartum experiences. These data allow researchers to examine population-level patterns related to maternal and infant health.

PRAMS Phase 8 includes several hundred variables capturing maternal demographics, pre-pregnancy health conditions, prenatal care experiences, health behaviors, insurance coverage, mental health indicators, and postpartum outcomes. The dataset contains a mix of categorical and numeric variables and reflects real-world survey complexity, including skip patterns and missing data.

Across participating states, PRAMS Phase 8 includes tens of thousands of respondents, providing sufficient sample size and variable richness to support exploratory analysis and regression modeling. For this project, variables related to prenatal mental health screening, pre-pregnancy mental health history, postpartum depressive symptoms, and sociodemographic characteristics were selected for analysis.

## Questions
The primary research question for this study is:

Are individuals who report being asked about depression or anxiety during prenatal care less likely to report postpartum depressive symptoms?

The primary outcome of interest is self-reported postpartum depressive symptoms, measured using PRAMS survey items assessing feelings of depression and loss of interest since birth. The primary predictors of interest include indicators of prenatal mental health screening, such as whether a healthcare provider asked about depression or anxiety during prenatal visits.

Additional covariates will include pre-pregnancy mental health history, insurance coverage, income, education, maternal age, and race or ethnicity.

# Methods 

## Use of Computational Assistance
Artificial intelligence tool (ChatGPT) was used to assist with code troubleshooting. All code and analytical decisions were reviewed, modified as needed, and implemented by the author. The author takes full responsibility for the accuracy, interpretation, and reproducibility of all analyses and results.

## Analysis Approach

At this stage of the project, the analysis plan includes descriptive summaries of key variables and exploratory comparisons between individuals who did and did not report prenatal mental health screening. These exploratory analyses will be used to assess data quality, missingness, and preliminary patterns in the data.

Later stages of the project will involve regression-based and machine learning approaches covered in the course to model associations between prenatal screening and postpartum mental health outcomes while accounting for relevant covariates. The richness of PRAMS survey content and linked birth data makes this analysis feasible within the scope of a semester-long project.

## Data Source

Data were obtained from the Pregnancy Risk Assessment Monitoring System (PRAMS), a population-based surveillance system conducted by the Centers for Disease Control and Prevention in collaboration with state health departments. PRAMS collects self-reported information on maternal experiences before, during, and shortly after pregnancy.

Because PRAMS uses a complex survey design, weighted analyses were conducted using the survey design variables provided in the dataset. Survey weights were applied to account for the sampling design, nonresponse, and noncoverage. Weighted descriptive estimates and weighted logistic regression were used to examine the association between prenatal mental health screening and postpartum depressive symptoms while adjusting for depression during pregnancy and maternal education.

## Study Population
The full PRAMS analytic dataset included `r scales::comma(n_total)` respondents. After excluding observations with missing values for the primary exposure, outcome, and selected covariates used in the main model, the final analytic sample included `r scales::comma(n_analysis)` respondents.


```{r}
#| label: tbl-study-pop
#| tbl-cap: "Study population characteristics"

study_pop_tbl <- tibble(
  Characteristic = c(
    "Sample size",
    "  Total respondents in full dataset",
    "  Respondents in analytic sample",
    "Prenatal screening",
    "  Screened during prenatal care",
    "  Not screened during prenatal care",
    "Postpartum depressive symptoms",
    "  Yes",
    "  No",
    "Depression during pregnancy",
    "  Yes",
    "  No",
    "Maternal education",
    "  Less than high school",
    "  High school graduate",
    "  Some college",
    "  College graduate",
    "  More than college"
  ),
  n = c(
    NA,
    n_total,
    n_analysis,
    NA,
    sum(analytic_dat$screened == 1, na.rm = TRUE),
    sum(analytic_dat$screened == 0, na.rm = TRUE),
    NA,
    sum(analytic_dat$pp_depress_bin == 1, na.rm = TRUE),
    sum(analytic_dat$pp_depress_bin == 0, na.rm = TRUE),
    NA,
    sum(analytic_dat$BPG_DEPRS8 == 1, na.rm = TRUE),
    sum(analytic_dat$BPG_DEPRS8 == 0, na.rm = TRUE),
    NA,
    sum(analytic_dat$MAT_ED == 1, na.rm = TRUE),
    sum(analytic_dat$MAT_ED == 2, na.rm = TRUE),
    sum(analytic_dat$MAT_ED == 3, na.rm = TRUE),
    sum(analytic_dat$MAT_ED == 4, na.rm = TRUE),
    sum(analytic_dat$MAT_ED == 5, na.rm = TRUE)
  )
) |>
  mutate(
    Percent = ifelse(is.na(n), NA, round(100 * n / n_total, 1)),
    n = ifelse(is.na(n), "", scales::comma(n)),
    Percent = ifelse(is.na(Percent), "", as.character(Percent))
  )

knitr::kable(study_pop_tbl, align = c("l", "r", "r"))

```

## Measures

The primary exposure in this study was self-reported prenatal mental health screening, measured using a PRAMS survey item indicating whether a healthcare provider asked about depression or anxiety during prenatal care. This variable was coded as a binary indicator representing whether respondents reported being screened during prenatal visits.

The primary outcome was postpartum depressive symptoms. PRAMS includes survey items assessing whether respondents experienced feelings consistent with depression or loss of interest following childbirth. These items have been used in previous research examining postpartum depressive symptoms using PRAMS data @bauman2020vitalsigns @farr2016provider.

Key covariates included prior depression during pregnancy (BPG_DEPRS8) and maternal education (MAT_ED). Prior depression during pregnancy was included as an important potential confounder, as individuals with a history of depression may be more likely to experience postpartum depressive symptoms. Maternal education was included as a sociodemographic variable that may influence both healthcare experiences and mental health outcomes.

Together, these variables allow for exploratory analysis of the relationship between prenatal mental health screening and postpartum depressive symptoms while accounting for selected demographic and clinical characteristics.

## Statistical Approach

Exploratory analyses included descriptive statistics, cross-tabulations, and graphical comparisons of postpartum depressive symptoms by screening status. All analyses were conducted in R.

Multivariable logistic regression models were specified with postpartum depressive symptoms as the binary outcome. Prenatal mental health screening was included as the primary exposure. Covariates included prior depression during pregnancy and maternal education. All categorical variables were coded with defined reference categories. Results are presented as log-odds estimates with corresponding confidence intervals.


## Results

The analytic sample included approximately 237,000 respondents after excluding missing outcome and exposure values. Overall, 81% of respondents reported being screened for depression during prenatal care, and 14% reported postpartum depressive symptoms.

```{r}
#| label: tbl-multivariable
#| tbl-cap: "Adjusted log odds for postpartum depressive symptoms"

multi_clean <- multi_tbl |>
  filter(term != "(Intercept)") |>
  mutate(
    Predictor = case_when(
      term == "screened1" ~ "Screened during prenatal care",
      term == "BPG_DEPRS82" ~ "Depression during pregnancy",
      term == "MAT_ED2" ~ "High school graduate",
      term == "MAT_ED3" ~ "Some college",
      term == "MAT_ED4" ~ "College graduate",
      term == "MAT_ED5" ~ "More than college",
      TRUE ~ term
    ),
    OR = exp(estimate),
    CI_low = exp(conf.low),
    CI_high = exp(conf.high)
  ) |>
  transmute(
    Predictor,
    `Odds ratio` = round(OR, 2),
    `95% CI` = paste0("(", round(CI_low, 2), ", ", round(CI_high, 2), ")"),
    `P value` = scales::pvalue(p.value)
  )

knitr::kable(multi_clean, align = "lccc")
```

```{r}
library(ggplot2)

fig_data <- analytic_dat |>
  group_by(screened) |>
  summarize(
    pct_depressed = mean(pp_depress_bin) * 100
  ) |>
  mutate(
    Screening = ifelse(screened == 1, "Screened", "Not screened")
  )

ggplot(fig_data, aes(x = Screening, y = pct_depressed)) +
  geom_col() +
  labs(
    title = "Postpartum depressive symptoms by prenatal screening",
    y = "Percent with depressive symptoms",
    x = ""
  ) +
  theme_minimal()
```

### Exploratory Analysis

We first examined the distribution of prenatal mental health screening and postpartum depressive symptoms in the analytic sample. We also explored how postpartum depressivmultie symptoms varied by screening status, prior depression during pregnancy, and maternal education.

Preliminary descriptive results suggested that postpartum depressive symptoms were more common among respondents who did not report receiving prenatal mental health screening. Higher rates of postpartum depressive symptoms were also observed among respondents who reported depression during pregnancy. Differences across maternal education groups were also examined to explore potential patterns in the data.

These exploratory analyses help identify initial patterns in the dataset and guide the selection of variables for later modeling and statistical analysis.

```{r}
#| label: fig-screening
#| fig-cap: "Postpartum depressive symptoms by prenatal mental health screening"

knitr::include_graphics(
  here("results","figures","screening_vs_postpartum_depression.png")
)
```

Individuals who reported receiving prenatal mental health screening had a lower prevalence of postpartum depressive symptoms (13.9% vs 15.2%). This corresponds to a 1.3 percentage point difference.

In adjusted analyses, prenatal mental health screening was associated with lower odds of postpartum depressive symptoms. Prior depression during pregnancy remained the strongest predictor of postpartum depressive symptoms, with substantially higher odds observed among individuals with a history of depression. Maternal education also showed a graded association with postpartum depressive symptoms.


```{r}
#| label: tbl-screening
#| tbl-cap: "Distribution of postpartum depressive symptoms by prenatal mental health screening"

library(dplyr)
library(tidyr)
library(scales)

screen_totals <- analytic_dat |>
  count(screened) |>
  mutate(
    screening = case_when(
      screened == 0 ~ "Not screened",
      screened == 1 ~ "Screened"
    ),
    Total = comma(n)
  ) |>
  select(screening, Total)

screening_table <- analytic_dat |>
  count(screened, pp_depress_bin) |>
  group_by(screened) |>
  mutate(
    percent = round(100 * n / sum(n), 1),
    Screening = ifelse(screened == 1, "Screened", "Not screened"),
    Outcome = ifelse(pp_depress_bin == 1, "Depressive symptoms", "No depressive symptoms")
  ) |>
  mutate(value = paste0(scales::comma(n), " (", percent, "%)")) |>
  select(Screening, Outcome, value) |>
  pivot_wider(names_from = Outcome, values_from = value)

knitr::kable(screening_table, caption = "Postpartum depressive symptoms by prenatal screening status")

```


Table 2 shows that among respondents who were not screened during prenatal care, 15.2% reported postpartum depressive symptoms. Among those who were screened, 13.9% reported postpartum depressive symptoms.

Logistic regression and random forest models demonstrated nearly identical performance, with comparable accuracy and ROC AUC values (Table 2). This suggests that more complex machine learning approaches did not substantially improve predictive performance relative to a simpler regression model.

In exploratory analyses, postpartum depressive symptoms were more common among women who did not report receiving prenatal depression screening compared to those who were screened. Prior depression during pregnancy was strongly associated with postpartum depressive symptoms, suggesting that mental health history is an important confounder to consider in adjusted analyses. Differences in postpartum depressive symptoms were also observed across education levels.

In the adjusted model, prenatal mental health screening was associated with higher odds of postpartum depressive symptoms (β = 0.759, 95% CI: 0.737–0.783). Prior depression during pregnancy was the strongest predictor (β = 3.376), indicating substantially higher odds of postpartum depressive symptoms among individuals with a history of depression. Maternal education also showed a graded association, suggesting potential socioeconomic differences in postpartum mental health outcomes.

## Discussion

This study evaluated the association between prenatal mental health screening and postpartum depressive symptoms using population-based PRAMS data. Results suggest that individuals who reported receiving prenatal screening had slightly lower rates of postpartum depressive symptoms, even after adjusting for key covariates.

These findings are generally consistent with existing research showing that postpartum depressive symptoms remain a significant public health concern in the United States. Previous analyses using PRAMS data have documented that a substantial proportion of women report postpartum depressive symptoms and that provider discussions about perinatal depression vary across populations and healthcare settings @bauman2020vitalsigns @farr2016provider. Clinical guidelines from organizations such as the American College of Obstetricians and Gynecologists recommend routine screening for depression during pregnancy and the postpartum period to support earlier identification and intervention @acog2023screening.

The exploratory results from this study suggest that prenatal screening may be associated with differences in postpartum mental health outcomes at the population level. However, these results should be interpreted cautiously. Because the data are observational and based on self-reported survey responses, the analyses cannot determine whether prenatal screening itself reduces postpartum depressive symptoms. It is possible that individuals who receive more comprehensive prenatal care are also more likely to receive screening and additional support services that influence postpartum mental health outcomes.

Prior depression during pregnancy emerged as one of the strongest predictors of postpartum depressive symptoms in the exploratory analyses. This finding aligns with previous research showing that mental health history is an important risk factor for postpartum depression. Accounting for prior depression is therefore essential when examining associations between prenatal care experiences and postpartum mental health outcomes.

Several strengths of this study should be noted. PRAMS provides a large, population-based dataset that captures maternal experiences across multiple states. The survey includes detailed information on prenatal care experiences, health behaviors, and postpartum outcomes, allowing researchers to examine mental health patterns at a population level. The large sample size also supports exploratory regression analyses and the evaluation of multiple potential confounders.

This study also has several limitations. First, PRAMS data are self-reported and may be subject to recall bias or reporting bias. Second, the postpartum depressive symptom measure used in PRAMS is not a clinical diagnosis of depression but rather a survey-based indicator of depressive symptoms. Third, the cross-sectional nature of the survey limits the ability to determine temporal relationships or causal effects. Finally, unmeasured factors such as healthcare access, provider practices, or social support may influence both screening and postpartum mental health outcomes.

Despite these limitations, the findings highlight the potential value of examining prenatal mental health screening practices using population-based surveillance data. Understanding how screening practices relate to maternal mental health outcomes may help inform future research and public health efforts aimed at improving maternal mental health.

Future analyses could expand on this work by incorporating additional covariates such as insurance status, socioeconomic indicators, and access to mental health services. More advanced modeling approaches, including machine learning techniques covered later in this course, could also be used to explore predictors of postpartum depressive symptoms and identify subgroups that may benefit most from screening and early intervention.

Overall, this study demonstrates how large surveillance datasets such as PRAMS can be used to explore maternal mental health patterns and generate hypotheses for future research. Continued attention to perinatal mental health remains important for improving maternal and infant health outcomes.

## References

