---
title: "04 Statistical Analysis"
author: "Alexandra Tejada-Strop"
format: html
execute:
  echo: true
  warning: false
  message: false
---

This page performs the main statistical analyses for the project.

It fits a bivariate logistic regression model examining the association between prenatal mental health screening and postpartum depressive symptoms, followed by a multivariable logistic regression adjusting for selected covariates. A tidymodels workflow is also fit and saved for downstream use.

```{r setup}
library(tidyverse)
library(here)
library(tidymodels)
library(broom)
```

```{r}
# Load processed analytic dataset
data_path <- here("data", "processed-data", "prams_analytic.rds")
stopifnot(file.exists(data_path))

dat <- readRDS(data_path)

# Create results folders if they do not already exist
dir.create(here("results", "tables"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("results", "output"), showWarnings = FALSE, recursive = TRUE)

# Quick check of imported data
glimpse(dat)
```

```{r}
# Define variables
outcome <- "pp_depress_bin"
main_exposure <- "screened"
covariates <- c("bpg_deprs8", "mat_ed")

all_vars <- c(outcome, main_exposure, covariates)
stopifnot(all(all_vars %in% names(dat)))

# Create complete-case analysis dataset
analysis_dat <- dat |>
  select(all_of(all_vars)) |>
  drop_na() |>
  mutate(
    screened = as.factor(screened),
    bpg_deprs8 = as.factor(bpg_deprs8),
    mat_ed = as.factor(mat_ed)
  )
```

# Check outcome coding before modeling
```{r}
table(analysis_dat$pp_depress_bin, useNA = "ifany")
```

# Save the complete-case dataset used for modeling
```{r}

saveRDS(
  analysis_dat,
  here("results", "output", "analysis_dataset_complete_case.rds")
)
```

# Basic analytic sample summary
```{r}
analysis_dat |>
  summarise(
    n = n(),
    postpartum_depression_cases = sum(pp_depress_bin == 1, na.rm = TRUE)
  )
```

# Bivariate logistic regression model
```{r}
fit_biv <- glm(
  pp_depress_bin ~ screened,
  data = analysis_dat,
  family = binomial()
)

biv_tbl <- tidy(fit_biv, conf.int = TRUE, exponentiate = TRUE)

biv_tbl
```

# Save bivariate results
```{r}

saveRDS(biv_tbl, here("results", "tables", "glm_bivariate_screened.rds"))
saveRDS(fit_biv, here("results", "output", "glm_bivariate_screened.rds"))
```

# Multivariable logistic regression model
```{r}
fit_multi <- glm(
  pp_depress_bin ~ screened + bpg_deprs8 + mat_ed,
  data = analysis_dat,
  family = binomial()
)

multi_tbl <- tidy(fit_multi, conf.int = TRUE, exponentiate = TRUE)

multi_tbl
```

# Save multivariable results
```{r}
saveRDS(multi_tbl, here("results", "tables", "glm_multivariable_main.rds"))
saveRDS(fit_multi, here("results", "output", "glm_multivariable_main.rds"))
```

# Tidymodels expects a factor outcome for classification
```{r}
analysis_dat2 <- analysis_dat |>
  mutate(pp_depress_bin = factor(pp_depress_bin, levels = c(0, 1)))

rec <- recipe(pp_depress_bin ~ screened + bpg_deprs8 + mat_ed, data = analysis_dat2) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

wf <- workflow() |>
  add_recipe(rec) |>
  add_model(logistic_reg() |> set_engine("glm"))

fit_wf <- fit(wf, data = analysis_dat2)
```

# Save tidymodels workflow
```{r}
saveRDS(fit_wf, here("results", "output", "tidymodels_logistic_main.rds"))

fit_wf
```

# Confirm output files were successfully saved
```{r}
file.exists(here("results", "output", "analysis_dataset_complete_case.rds"))
file.exists(here("results", "tables", "glm_bivariate_screened.rds"))
file.exists(here("results", "output", "glm_bivariate_screened.rds"))
file.exists(here("results", "tables", "glm_multivariable_main.rds"))
file.exists(here("results", "output", "glm_multivariable_main.rds"))
file.exists(here("results", "output", "tidymodels_logistic_main.rds"))
```

